---
title: "Compare DEG tools"
author: "Mark Ziemann"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
theme: cosmo
---

# Intro

Here the purpose is to demonstrate how the choice of DE tool can impact the results of a study.

Source code: Source code: https://github.com/markziemann/de_compare

View result: http://118.138.234.73/public/blog/deg_compare.html

## Libraries

```{r,libs}
library("edgeR")
library("DESeq2")
library("eulerr")
library("UpSetR")

```

## Input data

This is real data from an ATAC-seq experiment.
I have scrubbed the sample and gene names.

```{r,input}
x <- read.table("counts.mx",row.names=1,header=TRUE)
xx <- x[which(rowMeans(x)>10),]
str(xx)
dim(xx)
ss <- as.data.frame(colnames(xx))
ss$trt <- as.integer(grepl("T",ss[,1]))
rownames(ss) <- ss[,1]
ss[,1]=NULL
ss
```

## MDS analysis

This will give us an idea about which samples are similar/different.

```{r,mds,fig.height=7,fig.width=7}
plotMDS(xx, labels=rownames(ss),col=as.integer(as.factor(ss$trt)),main="MDS")
```

## EdgeR LRT

This was the method that was used in the group when I first got involved in this project.

```{r,edger}

design <- model.matrix(~ss$trt)
rownames(design) <- rownames(ss)
z <- DGEList(counts=xx)
z <- calcNormFactors(z)
z <- estimateDisp(z, design,robust=TRUE,prior.df=1)
fit <- glmFit(z, design)
lrt<-glmLRT(fit)
dge<-as.data.frame(topTags(lrt,n=Inf))
dge$dispersion<-lrt$dispersion
dge<-merge(dge,lrt$fitted.values,by='row.names')
rownames(dge)=dge$Row.names
dge$Row.names=NULL
dge<-dge[order(dge$PValue),]
head(dge,10)
dge_edger <- dge
sig <- subset(dge_edger,FDR<0.05)
dge_edger_up <- rownames(subset(sig,logFC>0))
dge_edger_dn <- rownames(subset(sig,logFC<0))
length(dge_edger_up)
length(dge_edger_dn)

```

## EdgeR QL

This is a "new" edgeR method which is more conservative.

```{r,edgerql}

design <- model.matrix(~ss$trt)
rownames(design) <- rownames(ss)
z <- DGEList(counts=xx)
z <- calcNormFactors(z)
z <- estimateDisp(z, design,robust=TRUE,prior.df=1)
fit <- glmQLFit(z, design)
lrt <- glmQLFTest(fit)
dge<-as.data.frame(topTags(lrt,n=Inf))
dge$dispersion<-lrt$dispersion
dge<-merge(dge,lrt$fitted.values,by='row.names')
rownames(dge)=dge$Row.names
dge$Row.names=NULL
dge<-dge[order(dge$PValue),]
head(dge,10)
dge_edgerql <- dge
sig <- subset(dge_edgerql,FDR<0.05)
dge_edgerql_up <- rownames(subset(sig,logFC>0))
dge_edgerql_dn <- rownames(subset(sig,logFC<0))
length(dge_edgerql_up)
length(dge_edgerql_dn)

```

## DESeq2

This is the method that I prefer using these days.

```{r,deseq2}

dds <- DESeqDataSetFromMatrix(countData = xx , colData = ss, design = ~ trt )
res <- DESeq(dds)
z<- results(res)
vsd <- vst(dds, blind=FALSE)
zz<-cbind(as.data.frame(z),assay(vsd))
dge<-as.data.frame(zz[order(zz$pvalue),])
head(dge,10)
dge_deseq2 <- dge
sig <- subset(dge,padj<0.05)
dge_deseq2_up <- rownames(subset(sig,log2FoldChange>0))
dge_deseq2_dn <- rownames(subset(sig,log2FoldChange<0))
length(dge_deseq2_up)
length(dge_deseq2_dn)

```



## Voom/Limma

This was suggested to me.

```{r,voom}

design <- model.matrix(~ss$trt)
v <- voom(xx, design, plot=FALSE)
vfit <- lmFit(v, design)
efit <- eBayes(vfit)
dge <- topTable(efit,n=Inf)
head(dge,10)
dge_voom <- dge
sig <- subset(dge_voom,adj.P.Val<0.05)
dge_voom_up <- rownames(subset(sig,logFC>0))
dge_voom_dn <- rownames(subset(sig,logFC<0))
length(dge_voom_up)
length(dge_voom_dn)

```

# Venn diagram

With the "eulerr" project.

```{r,venn}

v1 <- list("edgeR up"=dge_edger_up, "edgeR dn"=dge_edger_dn,
  "DESeq2 up"=dge_deseq2_up,"DESeq2 dn"=dge_deseq2_dn)
plot(euler(v1),quantities = TRUE)

v1 <- list("edgeR up"=dge_edger_up, "edgeR dn"=dge_edger_dn,
  "DESeq2 up"=dge_deseq2_up,"DESeq2 dn"=dge_deseq2_dn,
  "edgeR_QL up"=dge_edgerql_up, "edgeR_QL dn"=dge_edgerql_dn)
plot(euler(v1),quantities = TRUE)

v1 <- list("edgeR up"=dge_edger_up, "edgeR dn"=dge_edger_dn,
  "DESeq2 up"=dge_deseq2_up,"DESeq2 dn"=dge_deseq2_dn,
  "Voom up"=dge_voom_up, "Voom dn"=dge_voom_dn)
plot(euler(v1),quantities = TRUE)

```

# Upset plot

This should demonstrate the overlaps between different sets better.

```{r,upset}
v1 <- list("edgeR up"=dge_edger_up, 
  "DESeq2 up"=dge_deseq2_up,
  "Voom up"=dge_voom_up, 
  "edgeR_QL up"=dge_edgerql_up)

upset(fromList(v1), order.by = "freq")

v1 <- list("edgeR dn"=dge_edger_dn,
  "DESeq2 dn"=dge_deseq2_dn,
  "Voom dn"=dge_voom_dn,
  "edgeR_QL dn"=dge_edgerql_dn)

upset(fromList(v1), order.by = "freq")

```

## Session information

```{r,session}
sessionInfo()
```

